---
title: "Trends and Patterns among Grammy Award Winners (1965–2024)"
author: Deesha and Arina
output: github_document
---

```{r}
knitr::opts_chunk$set(
echo = FALSE, message = FALSE, warning = FALSE,
fig.align = "center", fig.retina = 2
)

library(tidyverse)
library(janitor)
set.seed(2025)
```


```{r}
list.files("data")
```


1. Introduction

The Grammy Awards are one of the most visible long-running signals of recognition in popular music and related genres.
This report explores how recognition patterns change over time (1965–2024) by comparing trends in nominations and (where available) wins, identifying repeat-recognized artists, and examining how the category landscape evolves across decades.

Roadmap:

describe the dataset and key variables,

explore major temporal patterns (growth, shifts, and “era” effects),

stress-test surprising patterns with sanity checks,

summarize takeaways and propose follow-up extensions.


2. Data
This report expects a CSV inside a data/ folder 


```{r}

data_dir <- "data"
csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)

if (length(csv_files) == 0) {
  stop(
    "No CSV found in ./data. Please add your Grammy dataset CSV to the data/ folder ",
    "and re-knit (e.g., data/grammy.csv)."
  )
}

raw <- readr::read_csv(csv_files[1], show_col_types = FALSE)

```


2.2 Variable discovery
Because Kaggle datasets sometimes use slightly different column names, this block tries to automatically map the key fields we need (year, category, nominee/artist, and winner flag if present).


```{r}
col <- names(raw)

pick <- function(pattern) {
match <- col[str_detect(col, regex(pattern, ignore_case = TRUE))]
if (length(match) == 0) return(NA_character_)
match[1]
}

cols <- list(
year     = pick("year"),
category = pick("category"),
nominee  = pick("nominee|artist|performer|group"),
winner   = pick("winner|win")   # may be NA if the dataset encodes wins differently
)

cols
```

2.3 Cleaned analysis table (standaerdized fields)

```{r}
df <- raw %>%
  mutate(
    year = if (!is.na(cols$year)) as.integer(.data[[cols$year]]) else NA_integer_,

    category = as.character(.data[["award"]]),

    nominee = if (!is.na(cols$nominee)) as.character(.data[[cols$nominee]]) else NA_character_,
    winner_raw = if (!is.na(cols$winner)) as.character(.data[[cols$winner]]) else NA_character_
  ) %>%
  mutate(
    winner = case_when(
      is.na(winner_raw) ~ NA,
      str_to_lower(winner_raw) %in% c("yes","y","true","t","1","winner") ~ TRUE,
      str_to_lower(winner_raw) %in% c("no","n","false","f","0") ~ FALSE,
      TRUE ~ NA
    )
  ) %>%
  filter(!is.na(year)) %>%
  arrange(year)

glimpse(df)

```

3. Main 


3.1 Question 1: How has overall Grammy activity changed over time?

```{r}
df %>%
count(year, name = "entries") %>%
ggplot(aes(year, entries)) +
geom_line() +
geom_point(size = 1) +
labs(
title = "Grammy dataset activity over time (1965–2024)",
x = "Year", y = "Count (nominations/entries)",
caption = "Interpretation: growth can reflect more categories, more nominees per category, or both."
)
```


3.2 Question 2: Who are the most frequently recognized artists?


```{r}
top_noms <- df %>%
filter(!is.na(nominee) & nominee != "") %>%
count(nominee, sort = TRUE) %>%
slice_max(n, n = 10)

top_wins <- df %>%
filter(!is.na(nominee) & nominee != "", !is.na(winner) & winner) %>%
count(nominee, sort = TRUE) %>%
slice_max(n, n = 10)

knitr::kable(top_noms, caption = "Top 10 most frequently appearing nominees (nominations/entries).")
```


```{r}
knitr::kable(top_wins, caption = "Top 10 most frequently winning nominees (where a winner flag is available).")
```

```{r}
ggplot(top_noms, aes(x = reorder(nominee, n), y = n)) +
geom_col() +
coord_flip() +
labs(
title = "Top 10 most frequently nominated artists",
x = "Artist / nominee",
y = "Number of nominations"
)

```


3.3 Question 3: Which Grammy categories dominate overall?


```{r}
cat_counts <- df %>%
  filter(!is.na(category) & category != "") %>%
  count(category, sort = TRUE) %>%
  slice_max(n, n = 12)

knitr::kable(
  cat_counts,
  caption = "Most common Grammy categories in the dataset (top 12)."
)

```


```{r}
ggplot(cat_counts, aes(x = reorder(category, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Most common Grammy categories (top 12)",
    x = "Category",
    y = "Number of entries"
  )
```

3.4 Question 4: How does category diversity change by decade?

```{r}
df %>%
  mutate(decade = 10 * (year %/% 10)) %>%
  distinct(decade, award) %>%              
  count(decade, name = "unique_award_types") %>% 
  ggplot(aes(decade, unique_award_types)) +
  geom_col() +
  labs(
    title = "Category diversity by decade (unique award types)",
    x = "Decade",
    y = "Number of unique award categories",
    caption = "Counts reflect how many different award types existed in each decade."
  )

```


4. Main — Skepticism (checking unintuitive patterns + alternative explanations)


4.1 Data-quality checks (do the implied rules actually hold?)



```{r}
sanity <- df %>%
group_by(year, category) %>%
summarise(
entries = n(),
winners_reported = sum(if_else(winner == TRUE, 1L, 0L), na.rm = TRUE),
.groups = "drop"
)

# Flag potentially inconsistent combinations (e.g., many winners for one year-category)

suspects <- sanity %>% filter(winners_reported > 1)

list(
year_range = range(df$year, na.rm = TRUE),
total_rows = nrow(df),
suspect_year_category = nrow(suspects)
)
```


4.2 Re-checking conclusions under alternative views (nominations-only vs wins-only)


```{r}
# If winner info is missing, this section still documents what is (and isn't) measurable.

if (all(is.na(df$winner))) {
cat("Winner information is not reliably available in the detected columns; conclusions focus on nominations/entries trends.")
} else {
view_win <- df %>% filter(winner == TRUE) %>% count(year, name = "wins")
view_nom <- df %>% count(year, name = "nominations")

ggplot() +
geom_line(data = view_nom, aes(year, nominations), alpha = 0.9) +
geom_line(data = view_win, aes(year, wins), alpha = 0.9) +
labs(
title = "Nominations vs. Wins over time (when winner flag is available)",
x = "Year", y = "Count",
caption = "Comparing the two views helps separate growth in nominations from growth in wins."
)
}
```


5. Organization (what we found, in the best order)

Across our four questions, we see a clear story about how the Grammys have
grown and changed. First, the number of entries per year has generally
increased over time, with especially fast growth after the 1980s, suggesting
both more categories and more nominees per category. Second, recognition is
somewhat concentrated: a small group of artists such as Willie Nelson and
Barbra Streisand accumulate many nominations, although newer artists like
Taylor Swift and Beyoncé also appear near the top.

On the category side, Song of the Year, Record of the Year, and Album of the
Year are among the most common award categories, but there is also a long tail
of specialized awards. The number of distinct award types per decade rises
from the early years to the 2010s, reflecting how the Grammys added new genres
and more specific categories. The dip in the 2020s is partly because we only
observe part of that decade so far.

6. Tables and charts (how to read them)

Key visual design decisions used across the report: consistent minimal theme, labeled axes, and captions that state the interpretation and the main caveat (e.g., “growth may reflect rule changes rather than purely more music”).

7. Conclusions (what we learned + next questions)

Overall, our analysis suggests that the Grammys have become larger and more
complex over time. The total volume of entries has grown, and the award
structure has expanded from a relatively small set of categories to hundreds
of distinct award types in recent decades. At the same time, recognition is
not evenly distributed: a small group of artists receive many nominations, and
some repeatedly appear among the top nominees and winners.

Our dataset has limitations. Winner information is incomplete in some rows,
category names sometimes change over time, and we do not observe details such
as genre tags or artist demographics. Future work could link this dataset to
external information like Billboard charts or streaming data to compare
industry recognition with commercial popularity, or to study how long new
categories survive once they are introduced.


8. Reproducibility (so anyone can rerun this end-to-end)


```{r}
required <- c("tidyverse", "janitor")
missing <- setdiff(required, rownames(installed.packages()))

if (length(missing) > 0) {
  cat(
    "Missing packages:", paste(missing, collapse = ", "), "\n",
    "Install them before knitting (e.g., install.packages(\"<pkg>\"))."
  )
} else {
  cat("All required packages installed.\n")
}

sessionInfo()

```



